Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Final State with Multivariate Techniques

### 2015-12-11 paper/sec-conclusions.tex

In this work we have presented a feasibility study for
 the measurement of Higgs pair production in the $b\bar{b}b\bar{b}$
final state at the High-Luminosity LHC with $\mathcal{L}=3$ ab$^{-1}$.
%
Our strategy was based on the combination of a traditional
cut-based analysis with state-of-the-art multivariate techniques.
%
We take into account 
all relevant backgrounds, in particular
QCD multijets, both  the irreducible $4b$
and the reducible 
$2b2j$ and $4j$ components.
%
We illustrate how the $2b2j$ component leads to
a contribution comparable to that of the $4b$ process,
due to a combination of  parton shower effects, $b$-quark 
pair radiation, and selection requirements.
%
Since any realistic study of the feasibility of Higgs pair
production at the HL-LHC should take into account the
high-PU environment in which such measurement will be performed,
in this work we have demonstrated the robustness of our analysis strategy
under high-PU conditions.

Combining the contributions from the resolved,
intermediate and boosted categories, we find that, for
$\mathcal{L}=3$ ab$^{-1}$, the
signal significance for
the production of Higgs pairs can be as large as $S/\sqrt{B}\simeq 4.0$.
%
This indicates that already from the $b\bar{b}b\bar{b}$
final state alone
it should be possible to claim evidence for Higgs pair production at
the HL-LHC.
%
Our study also suggests possible avenues that the LHC experiments
could explore to further improve this signal significance.
%
One handle could be to reduce the contribution from light and charm
jet mis-identification, ensuring that the
$2b2j$ component is subdominant over the $4b$ background.
%
Another possibility is to improve the mass resolution of the Higgs
reconstruction
in high-PU environments, and, more in general,
to optimize the PU subtraction
strategy, to reduce the impact of PU in the modelling
of kinematical variables, an effect that worsens the MVA discrimination.
%
Following these and related improvements in the experimental
analysis, it might be feasible to achieve
an observation of Higgs pair production even with only
$\mathcal{L}=300$ fb$^{-1}$.

One important implication of this work is that it should
be possible to improve  the accuracy on the extraction of
the Higgs trilinear coupling $\lambda$ from
a measurement of the
$\sigma\lp hh\to b\bar{b}b\bar{b}\rp$ cross-section, as compared
to existing estimates.
%
A determination of $\lambda$ in our approach is rather
non-trivial, involving
 not only regenerating signal samples
 for a wide range of values of  $\lambda$, but also
 repeating the analysis
optimisation, including the MVA training, for each
of these values.
%
This study is left to a future
publication, where we will also
compare with the corresponding  precision 
that has been obtained in other final states such as
 $b\bar{b}\gamma\gamma$
and $b\bar{b}\tau\tau$.
%
It will also be  interesting to perform
this exercise for a 100 TeV hadron collider~\cite{Barr:2014sga,
  Azatov:2015oxa,Papaefstathiou:2015iba,
  Arkani-Hamed:2015vfh}.
%
While signal yields will be certainly increased, also the (gluon-driven) QCD
multijet background will grow strongly.
%
Revisiting
the present analysis, including the MVA optimization,
at 100 TeV, will allow us
to assess the accuracy of an extraction of the trilinear
coupling $\lambda$ from the $b\bar{b}b\bar{b}$ final state
at 100 TeV.

Our analysis relies on the careful modeling of kinematical
distributions of signal and background events, since these provide
the inputs to the MVA discriminant.
%
In this respect, it would be important, having established the key
relevance of the $b\bar{b}b\bar{b}$ channel for the study of
Higgs pair production, to revisit and improve the
theoretical modeling of our signal and background simulation,
in particular using NLO calculations matched to
parton showers both for signal~\cite{Frederix:2014hta,Maierhofer:2013sha}
and for backgrounds~\cite{Alwall:2014hca,Gleisberg:2008ta}.
%

In this work we have considered only the SM production mechanism,
but many BSM scenarios predict deviations
in Higgs pair production, both at the level of total rates
and of
differential distributions.
%
In the absence of new explicit degrees of freedom,
deviations from the SM can be parametrized in
the EFT framework using higher-order
operators~\cite{Azatov:2015oxa,Goertz:2014qta}.
%
We thus plan to study which are the constraints
on the coefficients of these effective
operators that can be obtained from measurements
of various kinematical distributions
in the $hh\to b\bar{b}b\bar{b}$ process.
%
Note that the higher rates of the $b\bar{b}b\bar{b}$ final state as compared to
other final states, such as
$b\bar{b}\gamma\gamma$, allow to better constrain operators
that modify the high-energy behavior
of the theory, for instance,
it would become possible
to access the tail of the $m_{hh}$ distribution.

>>>>

In this work we have presented a feasibility study for
 the measurement of Higgs pair production in the $b\bar{b}b\bar{b}$
final state at the High-Luminosity LHC with $\mathcal{L}=3$ ab$^{-1}$.
%
Our strategy was based on the combination of a traditional
cut-based analysis with state-of-the-art multivariate techniques.
%
We take into account 
all relevant backgrounds, in particular
the irreducible $4b$
and the reducible 
$2b2j$ and $4j$ QCD multijets.
%
We illustrate how the $2b2j$ component leads to
a contribution comparable to that of the $4b$ process,
due to a combination of  parton shower effects, $b$-quark 
pair radiation, and selection requirements.
%
Since any realistic study of the feasibility of Higgs pair
production at the HL-LHC should take into account the
high-PU environment in which such measurement will be performed,
in this work we have demonstrated the robustness of our analysis strategy
under high-PU conditions.

Combining the contributions from the resolved,
intermediate and boosted categories, we find that, for
$\mathcal{L}=3$ ab$^{-1}$, the
signal significance for
the production of Higgs pairs can be as large as $S/\sqrt{B}\simeq 4.0$.
%
This indicates that already from the $b\bar{b}b\bar{b}$
final state alone
it should be possible to claim evidence for Higgs pair production at
the HL-LHC.
%
Our study also suggests possible avenues that the LHC experiments
could explore to further improve this signal significance.
%
One handle could be to reduce the contribution from light and charm
jet mis-identification, ensuring that the
$2b2j$ component is subdominant over the $4b$ background.
%
Another possibility would be to improve the mass resolution of the Higgs
reconstruction
in high-PU environments, and, more in general,
to optimize the PU subtraction
strategy, in order
to reduce the impact of PU in the modelling
of kinematical variables and the associated
degradation of the MVA discrimination.
%
Following these and related improvements in the experimental
analysis, it might be feasible to achieve
an observation of Higgs pair production even with only
$\mathcal{L}=300$ fb$^{-1}$.

One important implication of this work is that it should
be possible to improve  the accuracy on the extraction of
the Higgs trilinear coupling $\lambda$ from
a measurement of the
$\sigma\lp hh\to b\bar{b}b\bar{b}\rp$ cross-section, as compared
to existing estimates.
%
A determination of $\lambda$ in our approach is however
rather
non-trivial, involving
 not only regenerating signal samples
 for a wide range of values of  $\lambda$, but also
 repeating the analysis
optimisation, including the MVA training, for each
of these values.
%
This study is left to a future
publication, where we will also
compare with the corresponding  precision 
that has been reported from other final states such as
 $b\bar{b}\gamma\gamma$
and $b\bar{b}\tau\tau$.
%
It will also be  interesting to perform
this exercise for a 100 TeV hadron collider~\cite{Barr:2014sga,
  Azatov:2015oxa,Papaefstathiou:2015iba,
  Arkani-Hamed:2015vfh}.
%
While signal yields will be certainly increased, also the (gluon-driven) QCD
multijet background will grow strongly.
%
Revisiting
the present analysis, including the MVA optimization,
at 100 TeV, will allow us
to assess the accuracy of an extraction of the trilinear
coupling $\lambda$ from the $b\bar{b}b\bar{b}$ final state
at 100 TeV.

Our strategy relies on the modeling of the kinematical
distributions of signal and background events, since these provide
the inputs to the MVA discriminant.
%
In this respect, it would be important, having established the key
relevance of the $b\bar{b}b\bar{b}$ channel for the study of
Higgs pair production, to revisit and improve the
theoretical modeling of our signal and background simulation,
in particular using NLO calculations matched to
parton showers both for signal~\cite{Frederix:2014hta,Maierhofer:2013sha}
and for backgrounds~\cite{Alwall:2014hca,Gleisberg:2008ta}.
%

In this work we have considered only the SM production mechanism,
but many BSM scenarios predict deviations
in Higgs pair production, both at the level of total rates
and of
differential distributions.
%
In the absence of new explicit degrees of freedom,
deviations from the SM can be parametrized in
the EFT framework using higher-order
operators~\cite{Azatov:2015oxa,Goertz:2014qta}.
%
We thus plan to study the constraints
on the coefficients of these effective
operators that can be obtained from measurements
of various kinematical distributions
in the $hh\to b\bar{b}b\bar{b}$ process.
%
Note that the higher rates of the $b\bar{b}b\bar{b}$ final state as compared to
other final states, such as
$b\bar{b}\gamma\gamma$, allow to better constrain operators
that modify the high-energy behavior
of the theory, for instance,
it would become possible
to access the tail of the $m_{hh}$ distribution.

### 2015-12-11 paper/sec-mva.tex

In this section we study how the MVA results are modified when
signal and background events are embedded in a high-PU environment.
%
The simulation of PU and the corresponding subtraction with
{\tt SoftKiller} have been performed as described
in Sect.~\ref{sec:pileup}.
%
The cut-based analysis and the subsequent
MVA optimization have been performed using exactly the same
settings as in the case without PU.
%
In Table~\ref{table:cutflowMVA_PU} we provide the results
  at the end of the cut-based analysis,
for the case with $\la n_{\rm PU}\ra=80$+SK.
%
It is clear that the pre-MVA 
signal significance has been degraded
as compared to the case without PU.
%
We now find values for $S/\sqrt{B}$ of 0.25, 0.15 and 0.31, in the resolved,
intermediate and boosted categories, respectively, to be compared
with the corresponding values without PU, namely 1.0, 0.9 and 1.0,
see Table~\ref{table:cutflowMVA}. 
%

Subsequently to 
 the cut-based analysis, the surviving signal and background
MC events in the case of PU are used to re-train the MVA.
%
In Table~\ref{table:cutflowMVA_PU}
we also provide  the number of signal and
    background events expected
    at the HL-LHC with $\mathcal{L}=3$ ab$^{-1}$,
    as well as the
    corresponding values for $S/\sqrt{B}$ and $S/B$,
    for both the cut-based analysis (corresponding
    to $y_{\rm cut}=0$) and after the
    optimal MVA cut, for the simulations
    with $\la n_{\rm PU}\ra=80$
    and SK subtraction.
    %
    From Table~\ref{table:cutflowMVA_PU} we observe that, thanks
to the MVA, the signal significance and the
signal over background ratio are improved from 0.35 and $4\cdot 10^{-4}$
(0.25 and $1\cdot 10^{-4}$) up to 2.9 and 0.034 (2.5 and 0.15)
in the boosted (resolved) category.
%
The intermediate category, on the other hand, exhibits a rather lower post-MVA value of $S/\sqrt{B}$, around
1.1.
%
We thus conclude that, as was already the case
without PU,
the boosted category is the most promising
one for the study of Higgs pair production in the $b\bar{b}b\bar{b}$
final state
at the HL-LHC, also for the case of a high-PU environment, and that
important information will also be obtained from
the resolved category, which has a comparable significance.

@@

Another important result from Table~\ref{table:cutflowMVA_PU} is that,
for the optimal MVA cut, the signal over background ratio
is not unfeasibly small: we obtain that $S/B$ is 3.4\% (1.5\%)
in the boosted (resolved) categories.
%
This is indicates that while this measurement is still very challenging,
requiring a careful extraction from the data of the QCD
background, it is certainty within reach.
%
As in the case without PU, 
this analysis  indicates that a substantial increase in both
$S/B$ and $S/\sqrt{B}$ could be achieved by reducing the background component
that arises from the mis-identification of light and charm
fakes.

In Fig.~\ref{fig:nev2_PU}
we show the number of signal and background events that
are expected at the HL-LHC as a function of
$y_{\rm cut}$ and the corresponding ROC curve
%
Comparing Figs.~\ref{fig:nev2} and~\ref{fig:nev2_PU}, we observe
that the intermediate category is substantially degraded once PU effects
are accounted for, the reason being  sizable
a reduction in the number of signal
events that are now assigned to this category.
%
This shows that the selection criteria
of the intermediate category are less
resilient against PU contamination,
as opposed to the resolved and boosted selections.

@@

In Fig.~\ref{fig:sb_mva_PU} we show the signal significance,
$S/\sqrt{B}$, as well as the signal over background ratio,
$S/B$, including now the effects of PU.
%
The corresponding results in the case without PU were shown in
Fig.~\ref{fig:sb_mva}.
%
As can be seen, the MVA-driven enhancement is robust in the
presence of PU.
%
While unsurprisingly there us a degradation as compared to the case wo PU,
we still manage to achieve a signal significance of
around $S/\sqrt{B}$ between 2.5 and 3.0
separately for the boosted and resolved
categories.
%
On the other hand, the intermediate category becomes marginal.
%
We conclude that the qualitative conclusions drawn in
Sect.~\ref{sec:mva} are robust in the presence
of realistic PU effects.
%
Since no specific effort has been performed to
optimize PU subtraction, for example by tuning the value
of the patch length $a$ in {\tt SoftKiller}, we believe that
there is
still room for improvement in this respect.

@@

It is important to quantify which of the input variables
to the MVA carry the highest discrimination power
in the case of PU,
and compare it with the corresponding
results without PU.
%
We use the same estimator as in Sect.~\ref{sec:signalsignificance},
namely the sum
of the absolute value of all the weights connected to a given
input neuron $i$, Eq.~(\ref{eq:totweight}).
%
The results for the resolved and boosted categories are shown
on Fig.~\ref{fig:nnweights_PU}.
%
As we can see by comparing to the corresponding
results without PU in Fig.~\ref{fig:nnweights}, 
in the boosted category, the discrimination power of the invariant
mass of Higgs candidates is decreased and that of the various substructure
variables, in particular $C_2^{(\beta)}$ and
$D^{(\beta)}$, is conversely
increased.
%
This reflects the fact that these substructure variables are
relatively robust against PU contamination.
%
In addition, also the $p_T$ distributions of the AKT03
subjets within the large-$R$
jets carry important information.
%
In the case of the resolved category,  the highest
values of the ANN weights without PU
were found for the $p_T$ of the leading
Higgs and for the Higgs invariant masses.
%
This is also true in the case with PU, but now the rapidity difference
between the two Higgs candidates, $\Delta y_{hh}$ increases its
importance, again reflecting that this variable is relatively
insensitive to PU effects.
%

@@

We can finally combine the three categories, resolved,
intermediate and boosted, and determine the overall signal
significance, finding
\be
\lp \frac{S}{\sqrt{B}}\rp_{\rm tot} \simeq 4.0 \, ,
\ee
to be compared to the value $\lp S/\sqrt{B}\rp_{\rm tot} \simeq 8.5$
that was obtained in the case without PU.
%
We conclude that, even when realistic PU effects are accounted
for, at the HL-LHC the measurement of
Higgs pair production in the $b\bar{b}b\bar{b}$ final state should be 
well the threshold for evidence, and even meeting the
requirements for discovery might be within reach provided it is possible
to suppress the background from light and charm jet mis-identifications.

For the Run II luminosity, $\mathcal{L}=300$ fb$^{-1}$, one finds
instead $\lp S/\sqrt{B}\rp_{\rm tot} \simeq 1.3$, which increases up to
$\lp S/\sqrt{B}\rp_{\rm tot} \simeq 2.6$ if only the QCD $4b$ background is included.
%
This indicates that, interestingly, accessing Higgs pair production at Run II,
while still extremely challenging, does not seem to be completely impossible.
%
In addition, at Run II the average number of PU collisions is substantially
reduced as compared to the HL-LHC environment.
%
Therefore, it will be interesting to see if improvements in experimental $b$-jet
reconstruction and fake rejection at Run II are able to fulfill this
potential.

>>>>

In this section we study how the MVA results are modified when
signal and background events are embedded in a high-PU environment.
%
The simulation of PU and the corresponding subtraction with
{\tt SoftKiller} have been performed as described
in Sect.~\ref{sec:pileup}.
%
The cut-based analysis and the subsequent
MVA optimization have been performed using exactly the same
settings as in the case without PU.
%
In Table~\ref{table:cutflowMVA_PU} we provide the results
  at the end of the cut-based analysis,
  for the case with $\la n_{\rm PU}\ra=80$ with SK
  subtraction.
%
It is clear that the pre-MVA 
signal significance has been degraded
as compared to the case without PU.
%
We now find values for $S/\sqrt{B}$ of 0.25, 0.15 and 0.31, in the resolved,
intermediate and boosted categories, respectively, to be compared
with the corresponding values without PU, namely 1.0, 0.9 and 1.0,
see Table~\ref{table:cutflowMVA}. 
%

Subsequently to 
 the cut-based analysis, the signal and background
 MC events in the case of PU satisfying all selection cuts
  are used to re-train the MVA.
%
In Table~\ref{table:cutflowMVA_PU}
we also provide  the number of signal and
    background events expected
    at the HL-LHC with $\mathcal{L}=3$ ab$^{-1}$,
    as well as the
    corresponding values for $S/\sqrt{B}$ and $S/B$,
    for both the cut-based analysis (corresponding
    to $y_{\rm cut}=0$) and after the
    optimal MVA cut.
    %
    From Table~\ref{table:cutflowMVA_PU} we observe that, thanks
to the MVA, the signal significance and the
signal over background ratio is improved from 0.31 and $3\cdot 10^{-4}$
(0.25 and $10^{-4}$) up to 2.9 and 3.4\% (2.5 and 1.5\%)
in the boosted (resolved) category.
%
The intermediate category, on the other hand, exhibits a
significantly lower post-MVA value of $S/\sqrt{B}\simeq
1.1$.
%
We thus conclude that, as was already the case
without PU,
the boosted category is the most promising
one for the study of Higgs pair production in the $b\bar{b}b\bar{b}$
final state
at the HL-LHC, also for the case of a high-PU environment, but that
important information will also be obtained from
the resolved category, which has a comparable significance.

@@

Another important result from Table~\ref{table:cutflowMVA_PU} is that,
for the optimal MVA cut, the signal over background ratio
is not unfeasibly small: we obtain that $S/B$ is 3.4\% (1.5\%)
in the boosted (resolved) categories.
%
This indicates that while this measurement is still highly challenging,
requiring a careful extraction of the QCD
background from the data, it is certainty within reach.
%
As in the case without PU, 
our analysis  points out that both
$S/B$ and $S/\sqrt{B}$ could be increased  by reducing the background component
that arises from the mis-identification of light and charm
jets as $b$-jets.

In Fig.~\ref{fig:nev2_PU}
we show the number of signal and background events that
are expected at the HL-LHC as a function of
$y_{\rm cut}$, as well as the corresponding ROC curve.
%
Comparing Figs.~\ref{fig:nev2} and~\ref{fig:nev2_PU}, we observe
that the intermediate category is substantially degraded once PU effects
are accounted for due to a sizable
reduction in the number of signal
events that are now classified in this category.
%
This shows that the selection criteria
of the intermediate category are less
resilient against PU contamination,
as opposed to the resolved and boosted selections.

@@

In Fig.~\ref{fig:sb_mva_PU} we show the signal significance,
$S/\sqrt{B}$, and the signal over background ratio,
$S/B$, accounting now for the effects of PU.
%
The corresponding results in the case without PU were shown in
Fig.~\ref{fig:sb_mva}.
%
As can be seen, the MVA-driven enhancement is robust in the
presence of PU.
%
While as expected there is some degradation as compared to the case
without PU,
it is still possible to
achieve a signal significance of
around $S/\sqrt{B}$ between 2.5 and 3.0 (depending on the
specific choice of $y_{\rm cut}$)
separately for the boosted and resolved
categories.
%
On the other hand, the intermediate category now becomes marginal.
%
We conclude that the qualitative conclusions drawn
in the case without PU also hold when the analysis
is performed in a high-PU environment.
%
Since no specific effort has been performed to
optimize PU subtraction, for instance by tuning the value
of the patch length $a$ in {\tt SoftKiller}, we believe that
there could be still room for further  improvement.

@@

It is useful to quantify which of the input variables
to the MVA carry the highest discrimination power
in the case of PU,
and compare this with the corresponding
results without PU, by means of
Eq.~(\ref{eq:totweight}).
%
The 
results for the resolved and boosted categories are shown
on Fig.~\ref{fig:nnweights_PU}.
%
As we can see by comparing with Fig.~\ref{fig:nnweights}, 
in the boosted category, the discrimination power of the invariant
mass of Higgs candidates is decreased, while
that of  substructure
variables such as $C_2^{(\beta)}$ and
$D^{(\beta)}_2$ is conversely
increased.
%
In addition, also the $p_T$ distributions of the AKT03
subjets within the large-$R$
jets carry important information.
%
In the case of the resolved category,  the highest
values of the ANN weights without PU
were found for the $p_T$ of the 
Higgs candidates and their invariant masses, as well
as the for invariant mass of the di-Higgs system
and the $p_T$ of the small-$R$ jets.
%
As we can see in Fig.~\ref{fig:nnweights_PU},
these results  are reasonably stable in the case with PU.

@@

We can finally combine the three categories, resolved,
intermediate and boosted, and determine the overall signal
significance, finding
\be
\lp \frac{S}{\sqrt{B}}\rp_{\rm tot} \simeq 4.0 \, ,\quad
\mathcal{L}=3\,{\rm ab}^{-1}\, .
\ee
%
We thus conclude that the measurement of
Higgs pair production in the $b\bar{b}b\bar{b}$ final state
should be 
well above the threshold for evidence, and even meeting the
requirements for discovery might be within reach, provided it is possible
to suppress the background from light and charm jet mis-identifications.
%
Extrapolating the above results
to total integrated luminosity expected at the end of
LHC Run II, one finds
instead
\be
\lp \frac{S}{\sqrt{B}}\rp_{\rm tot} \simeq 1.3(2.6) \, ,\quad
\mathcal{L}=300\,{\rm fb}^{-1}\, ,
\ee
where the number in brackets corresponds to neglecting all backgrounds
except for the QCD $4b$ multijets.
%
This suggest that, remarkably, accessing Higgs pair production at Run II
might not be completely impossible, specially accounting
for the fact that the mean number of PU events per
bunch crossing is much smaller
that at the HL-LHC.
